---
title: "Exploratory Data Analysis of White Wine Quality"
author: "Leighann Irving"
date: "July 29, 2016"
output: html_document
---

```{r, echo=FALSE}
setwd('/Users/lirving/Documents/R Prac/EDA_Course_Materials/Final Project')
library(ggplot2)
library(gridExtra)
library(GGally)
white_wine <- read.csv("wineQualityWhites.csv")

white_wine <- subset(white_wine, select = -X) #Remove `X` since it is just a count of the observations and is not included in the attributes listed in the text file describing the data.
```


###Overview of the data
```{r}

str(white_wine)
summary(white_wine)
```

<!-- Note: The variables `fixed.acidity`, `volatile.acidity`,`citric.acid`, `residual.sugar`, `chlorides`, and `sulphates` were recorded in g/ dm^3^ units but since $1g/ dm^{3} = 1 litre$ I will describe these variables in litres. -->

Given the data set I want to find which variables influence the wine experts' ranking of `quality`.

Since `quality` must be an integer value between 0 and 10 (i.e. discrete and categorical) it should should be changed to a factor.

```{r}
white_wine$quality <- factor(white_wine$quality, ordered = TRUE)
```

##Univariate Analysis

###Looking at distribution of quality
```{r, message=FALSE}
qplot(quality, data = white_wine)
```

The `quality` of the white wine is approximately normally distributed. 

Interestingly, no white wine was determined to be "very excellent" and given a perfect score of 10/10. Contrastingly, none was determined to be "very bad" and given a score of 0, as well as a score of 1 or 2.

```{r}
table(white_wine$quality)
```

About 45% of the wines were given a score of 6 and 93% were between 5 and 7 therefore most were considered "average" by the wine experts.

###Histogram of each variable
```{r, message=FALSE}

fixed.acid.count <- qplot(fixed.acidity, data = white_wine)
vol.acid.count <- qplot(volatile.acidity, data = white_wine)
citric.acid.count <- qplot(citric.acid, data = white_wine)
res.sugar.count <- qplot(residual.sugar, data = white_wine)
chlorides.count <- qplot(chlorides, data = white_wine)
free.s02.count <- qplot(free.sulfur.dioxide, data = white_wine)
total.so2.count <- qplot(total.sulfur.dioxide, data = white_wine)
density.count <- qplot(density, data = white_wine)
pH.count <- qplot(pH, data = white_wine)
sulphates.count <- qplot(sulphates, data = white_wine)
alcohol.count <- qplot(alcohol, data = white_wine)

grid.arrange(fixed.acid.count, vol.acid.count, citric.acid.count, 
             res.sugar.count, chlorides.count, free.s02.count, total.so2.count,
             density.count, pH.count, sulphates.count, alcohol.count, ncol = 3)
```



###Closer look at residual.sugar

```{r, message=FALSE, warning=FALSE}
qplot(residual.sugar, data = white_wine, xlim = c(0,20), 
      color = I('black'), fill = I('purple'))
```

There is an extremely high peak between 1 and 2g / litre of `residual.sugar`.

```{r}
summary(white_wine$residual.sugar)
```

Since 45 g/ litre is the standard of sweetness, the low median (5.2 g/ litre) means that majority of the white wines were very bitter. Maybe the `quality` of the wines would have been higher on average had there been a larger percentage of sweet wines. This warrants an investigation of the relationship between `residual.sugar` and `quality`.


###Closer look at percent alcohol content
```{r, message=FALSE}
qplot(alcohol, data = white_wine, color = I('black'), fill = I('purple'))
```

The distribution is skewed right with a peak between 9 and 10% alcohol.

It looks like there are about three peaks in this distribution.

####Close-up of the first peak
```{r, message=FALSE, warning=FALSE}
qplot(alcohol, data = white_wine, color = I('black'), fill = I('purple'),
      xlim = c(8,10))
```

####Close-up of the second peak
```{r, message=FALSE, warning=FALSE}
qplot(alcohol, data = white_wine, color = I('black'), fill = I('purple'),
      xlim = c(10,12))
```

####Close-up of the third peak
```{r, message=FALSE, warning=FALSE}
qplot(alcohol, data = white_wine, color = I('black'), fill = I('purple'),
      xlim = c(12,14))
```

```{r, message=FALSE, warning=FALSE}
qplot(alcohol, data = white_wine, color = I('black'), fill = I('purple'),
      binwidth = 0.05)
```

From the close-ups and the more detailed histogram above it can be seen that there are evenly-spaced gaps (or `alcohol` levels with extremely low counts) among the distributions. Why would this happen?

```{r}
head(white_wine$alcohol,15)
tail(white_wine$alcohol,15)
```

Looking at some of the raw values of `alcohol`, the pattern in the distributions may be explained by most of the values rounded to one decimal place while a few (the gaps) were rounded to two.

###Closer look at volatile acidity

```{r, message=FALSE}
qplot(volatile.acidity, data = white_wine, color = I('black'), fill = I('purple'))
```

Most of the white wines had a `volatile acidity` around 0.25 g/ litre. The distribution appears somewhat symmetric but with a long right-tail. Why is this? "Too high" levels are said to lead to an unpleasant, vinegar taste so presumably this long tail has a negative relationship with quality. I will investigate this `volatile acidity`-`quality` relationship later on.

###Closer look at citric acid quantities

```{r, message=FALSE}
qplot(citric.acid, data = white_wine, color = I('black'), fill = I('purple'))
```

Between 0 and about 0.75g /litre, the histogram appears to be normally distributed. However, there is a long right-tail. This may be accounted for by the fact that citric acid is said to add 'freshness' and flavor to wines and as such more wines would be made with a high citric acid content. I will investigate the relationship between `citric acid` and `quality` later on.

###Closer look at potassium sulphates

```{r, message=FALSE}
qplot(sulphates, data = white_wine, color = I('black'), fill = I('purple'))
```

Most sulphates are between 0.3 and 0.75 litres.

```{r, message=FALSE}
qplot(sulphates, data = white_wine, color = I('black'), fill = I('purple'),
      binwidth = 0.01)
```

The histogram is approximately normally distributed but with a long-right tail.

###Closer look at free sulfur dioxide

```{r, message=FALSE}
qplot(free.sulfur.dioxide, data = white_wine, 
      color = I('black'), fill = I('purple'))
```

```{r, warning=FALSE}
qplot(free.sulfur.dioxide, data = white_wine, 
      color = I('black'), fill = I('purple'),
      xlim = c(0,100), binwidth = 1 )
```

It is said that at free sulfur dioxide (SO2) concentrations over 50g/ litre, SO2 becomes evident in the nose and taste of wine but there is no change in the pattern of the distribution around 50g / litre. 

However, there is a long right-tail so wine makers are making wine with disproportionately high free SO2 levels. Presumably this happens because the effect of high free SO2 concentrations is more pleasant to buyers. I shall test this by later examining the relationship between free SO2 levels and quality.


###Summary of Univariate Data:

There are 4,898 white wines in the dataset with 12 features(fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, quality).

The goal of this EDA is to ascertain which variables are good predictors of quality.

volatile.acidity, citric.acid, residual.sugar and free.sulfur.dioxide are the variables expected to be the best predictors.

##Bivariate Analysis

###Overview
```{r, message=FALSE}
set.seed(1234)
ggpairs(white_wine[sample.int(nrow(white_wine), 1000), ])
```

####Quality vs. Residual Sugar

```{r}
ggplot(aes(x = quality, y = residual.sugar), data = white_wine) + 
  geom_point(alpha = 1/4)
```


```{r}
plot(x = white_wine$quality, y = white_wine$residual.sugar)
```


```{r}
by(white_wine$residual.sugar, white_wine$quality, summary)
```


From the boxplot above it can be seen that the highest-rated white wines (`quality` ranking 9) had the smallest range and the lowest median of residual sugar however there seems to be no clear relationship between `quality` and `residual.sugar`.


####Quality vs. Citric Acid
```{r}
plot(x = white_wine$quality, y = white_wine$citric.acid)
```

From the boxplot above it can be seen that the median `citric.acid` are quite close among the `quality` scores and that the highest-rated white wines (`quality` ranking 9) had the highest median `citric.acid`. However, there seems to be no clear relationship between `quality` and `citric.acid`.

####Quality vs. Fee Sulfur Dioxide
```{r}
plot(x = white_wine$quality, y = white_wine$free.sulfur.dioxide)
by(white_wine$free.sulfur.dioxide, white_wine$quality, summary)
```

From the boxplot above it can be seen that the median `free.sulfur.dioxide` are quite close among the `quality` scores and there seems to be no clear relationship between the variables.

####Quality vs. pH
```{r}
plot(x = white_wine$quality, y = white_wine$pH)
```

From the boxplot above it can be seen that the highest-rated wines had the highest `pH` and from a quality ranking of 5 through 9 there is a corresponding increase in `pH` however it is hard to tell from just the plot whether the two variables are correlated therefore I shall conduct a correlation test.

```{r}
white_wine$quality <- as.numeric(white_wine$quality)
cor.test(white_wine$pH, white_wine$quality)
```

With a correlation coefficient of 0.10, no linear relationship exists between `pH` and `quality`.

```{r, echo=FALSE}
white_wine$quality <- factor(white_wine$quality)
```


####Quality vs. Volatile Acidity
```{r}
plot(x = white_wine$quality, y = white_wine$volatile.acidity)
by(white_wine$volatile.acidity, white_wine$quality, summary)
```

There doesn't seem to be a strong correlation between `volatile acidity` and `quality`.  

####Quality vs. Chlorides
```{r}
plot(x = white_wine$quality, y = white_wine$chlorides)
by(white_wine$chlorides, white_wine$quality, summary)
```

From the boxplot it can be seen that the median `chlorides` are quite close among the `quality` scores but it seems there could be a negative correlation between `chlorides` and `quality` with the highest ranked wines having the lowest amount of `chlorides` however to be sure I shall conduct a correlation test.

```{r}
white_wine$quality <- as.numeric(white_wine$quality)
cor.test(white_wine$chlorides, white_wine$quality)
```

With a correlation of -0.21, no linear relationship exists between `chlorides` and `quality`.

```{r, echo=FALSE}
white_wine$quality <- factor(white_wine$quality)
```


####Quality vs. Alcohol
```{r}
plot(x = white_wine$quality, y = white_wine$alcohol)
```

From the boxplot, the highest-rated wines had the highest median `alcohol` percentage and  `quality` scores 1-3 are correlated with decreasing `alcohol` while 3-9 are correlated with increasing `alcohol` therefore I shall conduct a correlation test to determine if there is an overall correlation.

```{r}
white_wine$quality <- as.numeric(white_wine$quality)
cor.test(white_wine$alcohol, white_wine$quality)
```

With a correlation coefficient of 0.44 there is a weak positive relationship between `alcohol` and `quality`.

```{r}
ggplot(aes (x = quality, y = alcohol), data = white_wine) +
  geom_point() + geom_smooth(method = 'lm', color = 'purple')
```


```{r, echo=FALSE}
white_wine$quality <- factor(white_wine$quality)
```

####Quality vs Density

```{r}
plot(x = white_wine$quality, y = white_wine$density)
```

From the boxplot the highest-rated wines had the lowest median `density` and it seems there may be a negative relationship between `density` and `quality`. To be sure I shall conduct a correlation test. 

```{r}
white_wine$quality <- as.numeric(white_wine$quality)
cor.test(white_wine$quality, white_wine$density)
```

With a correlation of -0.31 there is, in fact, a strong negative (linear) relationship between `density` and `quality`.

###Summary of Bivariate Data:

* No variable was strongly or moderately correlated with `quality`.
* There is a weak positive correlation between `alcohol` and `quality`.
* There is a weak negative correlation between `density` and `quality`.

```{r, echo=FALSE}
white_wine$quality <- factor(white_wine$quality)
```

###Multivariate Analysis

Since `alcohol` and `density` both affect `quality` let's plot `alcohol` against `density` distinguished by `quality` to see if there is a pattern.

####Alcohol vs Density by Quality
```{r}
ggplot(aes(x = alcohol, y = density), data = white_wine) +
  geom_point(aes(color = quality), stat = "summary", fun.y = median) 
```

There is a strong negative relationship between `alcohol` and `density` however, the `quality` scores are scattered among the plot and there is no pattern.

##Reflection

* None of the variables that were expected to influence `quality` (`residual.sugar`, `citric.acid`, `volatile.acidity` and `free.sulfur.dioxide`) did.
* Only 2 variables correlated with `quality` (`alcohol` and `density`) and even those were weak correlations.
* This indicates that the wine experts may determine `quality` based on some unknown variable, for example price.

